---
title: "join multiple metadata tibbles to one tibble"
output: html_document
date: "2025-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# combine dataset level infor
## examine a few sample meta
```{r}
df <- read.csv("output/human/GSE70630/GSE70630_dataset_metadata.csv")
tb <- read_csv("output/human/GSE70630/GSE70630_dataset_metadata.csv")
```


## combine human dataset meta
```{r }
# loop through the directory output/human/GSE*
human_dataset_lists <- list.files(
  path = "output/human",
  pattern = "^GSE\\d+_dataset_metadata\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)

# read in the file join the tibbles together
human_dataset_combined <- human_dataset_lists %>%
  set_names() %>%
  map_df(~ read_csv(.x, col_types = cols(.default = "c")), .id = "source_file")%>%
  select(-source_file)

# write the joined tibble to output/human/joined
write_csv(human_dataset_combined, file.path("output", "human_dataset_combined.csv"))
```

## combine mouse dataset meta
```{r }
# loop through the directory output/human/GSE*
mouse_dataset_lists <- list.files(
  path = "output/mouse",
  pattern = "^GSE\\d+_dataset_metadata\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)

# read in the file join the tibbles together
mouse_dataset_combined <- mouse_dataset_lists %>%
  set_names() %>%
  map_df(~ read_csv(.x, col_types = cols(.default = "c")), .id = "source_file")%>%
  select(-source_file)

# write the joined tibble to output
write_csv(mouse_dataset_combined, file.path("output", "mouse_dataset_combined.csv"))
```

# combine sample level infor
## examine a few sample meta
```{r}
gse70630 <- read_csv("output/human/GSE70630/GSE70630_sample_metadata.csv")
gse76381 <- read_csv("output/human/GSE76381/GSE76381_sample_metadata.csv")
gse84465 <- read_csv("output/human/GSE84465/GSE84465_sample_metadata.csv")
gse89567 <- read_csv("output/human/GSE89567/GSE89567_sample_metadata.csv")
gse97930 <- read_csv("output/human/GSE97930/GSE97930_sample_metadata.csv")
```


## combine human sample meta
```{r}
# Step 1: List the sample metadata files
human_sample_list <- list.files(
  path = "output/human", 
  pattern = "^GSE\\d+_sample_metadata\\.csv$", 
  recursive = TRUE, 
  full.names = TRUE
)

# Step 2: Combine using map_dfr which handles different columns
human_sample_combined <- human_sample_list %>%
  set_names() %>%
  map_dfr(~ read_csv(.x, col_types = cols(.default = "c")), .id = "source_file") %>%
  select(-source_file)

# Step 3: write the joined tibble to output
write_csv(human_sample_combined, file.path("output", "human_sample_combined.csv"))
```

## combine mouse sample meta
```{r}
# Step 1: List the sample metadata files
mouse_sample_list <- list.files(
  path = "output/mouse", 
  pattern = "^GSE\\d+_sample_metadata\\.csv$", 
  recursive = TRUE, 
  full.names = TRUE
)

# Step 2: Combine using map_dfr which handles different columns
mouse_sample_combined <- mouse_sample_list %>%
  set_names() %>%
  map_dfr(~ read_csv(.x, col_types = cols(.default = "c")), .id = "source_file") %>%
  select(-source_file)

# Step 3: write the joined tibble to output
write_csv(mouse_sample_combined, file.path("output", "mouse_sample_combined.csv"))
```

## reduce dimentionality of combined sample table
```{r}
# examine column names of human_sample_combined
human_sample_combined_colname <- colnames(human_sample_combined)

# examine selected columns
selected_col <- "tissue"

# get non NA for selected columns
human_sample_combined %>%
  filter(!is.na(.data[[selected_col]]) & .data[[selected_col]] != "") %>%
  pull(.data[[selected_col]])

# Calculate proportions
human_sample_combined %>%
  summarise(
    total = n(),
    non_missing = sum(!is.na(.data[[selected_col]]) & .data[[selected_col]] != ""),
    missing = sum(is.na(.data[[selected_col]]) | .data[[selected_col]] == ""),
    non_missing_pct = non_missing / total * 100,
    missing_pct = missing / total * 100
  )
```

### preselection
```{r}
# Define the columns to remove
columns_to_remove <- c(
  "cloneid", "protocol", "brain_bank",
  "braak_stage_(ad_neuropathological_progression)", 
  "braak_tangle_stage", "source", "culture_condition", "organoid_generation"
)
# Remove
human_sample_combined <- human_sample_combined %>%
  select(-any_of(columns_to_remove))
```

### reduce dimentionality of human_sample_combined
```{r}
# Collapse helper function
collapse_columns <- function(df, keywords, new_name) {
  cols_to_merge <- df %>%
    select(matches(paste(keywords, collapse = "|"), ignore.case = TRUE)) %>%
    colnames()

  message("Merging into ", new_name, ": ", paste(cols_to_merge, collapse = ", "))

  df %>%
    mutate(!!new_name := coalesce(!!!syms(cols_to_merge))) %>%
    select(-all_of(setdiff(cols_to_merge, new_name)))
}

# Apply to key features
human_sample_combined_reduced <- human_sample_combined %>%
  collapse_columns(c("age", "age_at_death", "age_\\(weeks\\)", "age_\\(years\\)", "age_\\(yr\\)"), "Age") %>%
  collapse_columns(c("sex", "gender"), "Sex") %>%
  collapse_columns(c("disease", "disease_state", "health_condition"), "Disease_status") %>%
  collapse_columns(c("tissue", "tissue_region", "culture_type", "sample_type", "region", "brain_region"), "Tissue") %>%
  # collapse_columns(c("developmental_stage", "development_stage", "stage"), "Stage") %>%
  collapse_columns(c("supplementary_file"), "Supplementary")
```

### examine the reduced sample meta
```{r}
# examine selected columns
selected_col_reduced <- "Supplementary"

# get non NA for selected columns
human_sample_combined_reduced %>%
  filter(!is.na(.data[[selected_col_reduced]]) & .data[[selected_col_reduced]] != "") %>%
  pull(.data[[selected_col_reduced]])

# Calculate proportions
human_sample_combined_reduced %>%
  summarise(
    total = n(),
    non_missing = sum(!is.na(.data[[selected_col_reduced]]) & .data[[selected_col_reduced]] != ""),
    missing = sum(is.na(.data[[selected_col_reduced]]) | .data[[selected_col_reduced]] == ""),
    non_missing_pct = non_missing / total * 100,
    missing_pct = missing / total * 100
  )
```


### visualize Missingness / Sparsity
```{r}
library(naniar)    # for missing value visualization
library(ggplot2)

# Visualize missingness heatmap on reduced metadata
gg_miss_upset(human_sample_combined_reduced)  # Summary of intersection of missingness (like Venn diagram)

# For a full heatmap (sample × feature)
human_sample_combined_reduced %>%
  select(-gse_id, -gsm_id) %>%
  mutate(across(everything(), ~ifelse(is.na(.), 0, 1))) %>%
  as.matrix() %>%
  image(
    main = "Missingness Heatmap",
    xlab = "Features",
    ylab = "Samples",
    col = c("white", "black")
  )
```



















